Long--term perspectives of social inequality in mortality among adults and elderly in northern Sweden 1851--2013\footnote{Session "Health, morbidity, and mortality" at the Nordic Demographic Symposium 14--16 June 2017, Turku, Finland}

### 2017-06-12 abo.Rnw
Small revision
Åbo version

To investigate the association between social class and mortality among adults and elderly from the 19th century to the present and how it is connected to socio-economic inequality in society.
This is analysed according to 
\begin{itemize}
\item gender
\item age group (working age vs retired)
\item cause of death (cardiovascular, cancer)
\item (education 1990--2013)
\item (income 1986-2005)

@@

\begin{frame}{Economic inequality}

@@

\begin{frame}{Outline}

\begin{center}
\begin{enumerate}

\item \emp{Data} management
\item \emp{Modeling} considerations
\item \emp{Analysis}
\item \emp{Results}
\item \emp{Implications} of the results
\end{enumerate}

\end{center}

@@

Remote area in northern Sweden, dominated by agriculture until well into the 20th century. 19th century communications were rudimentary.

@@

\item Small town, industrialization of the region during the 20th century.

@@

\begin{frame}{The sampling frame}

@@

\item[sex]
\item[birth date] By quarter (15 Feb, May, Aug, Nov)
\item[death date] By day (\emp{Socialstyrelsen})
\item[cause of death] from \emp{Socialstyrelsen}
\item[parish] Urban/Rural
\item[Social class] Elite, middle, worker. Household definition 1851--1950. Individual definition 1960--2013
\item[income] 1990--2005, grouped by quartiles.
\item[education] 1990--2005, three categories.
\item[period]  1851--1875, \ldots,  2008--2013 (10 periods). No results presented for women 1961-70, 1971-80 due to missing information on class.
%\item[cohort] 1761--1771, \ldots, 1971--1974 (22 birth cohorts).

@@

\begin{frame}[fragile]{Education 1990--2013}

<<educ>>=

educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
educ$period <- factor(educ$period)
educ <- educ[!is.na(educ$educ), ]
levels(educ$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
educ$educ <- relevel(educ$educ, ref = "middle")
educ$civst <- relevel(educ$civst, ref = "married")
fitm <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
             data = educ[educ$sex == "male", ])
fitf <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
             data = educ[educ$sex == "female", ])
cols <- c("darkgreen", "red", "black")
@

<<ploteduc>>=
oldpar <- par(mfrow = c(1, 2))

plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
axis(1)
axis(2, at = c(0, 0.5, 1.0), las = 1)
abline(h = 0.5, lty = 3)
box()
legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)

plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
axis(1)
axis(2, at = c(0, 0.5, 1.0), las = 1)
abline(h = 0.5, lty = 3)
box()
legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)

par(oldpar)
@

\begin{frame}{Income 1986--2005}

<<femeduc>>=
income <- lisaold[!is.na(lisaold$income), ]
income$period <- factor(income$period)
##income <- income[!is.na(income$income), ]
##levels(income$income) <- c("low", "low", "middle", "middle", "middle", "high", "high")
income$income <- relevel(income$income, ref = "low")
income$civst <- relevel(income$civst, ref = "married")
fitm <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
             data = income[income$sex == "male", ])
fitf <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
             data = income[income$sex == "female", ])
cols <- c("darkgreen", "red", "blue", "black")

<<plotincome>>=
oldpar <- par(mfrow = c(1, 2))

plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
axis(1)
axis(2, at = c(0, 0.5, 1.0), las = 1)
abline(h = 0.5, lty = 3)
box()
legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)

plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
axis(1)
axis(2, at = c(0, 0.5, 1.0), las = 1)
abline(h = 0.5, lty = 3)
box()
legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)

par(oldpar)
@

@@

\begin{itemize}
\item \emp{Improved} survival characterised all social classes.
\item \emp{Late} appearance of the “modern” pattern of social differences in mortality among men
\item Distinct \emp{gender} differences in the development.
\item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
\item No \emp{qualitative} but partly \emp{quantitative} differences between those in working age compared to the retired population.
\item \emp{Educational} and \emp{income} differences during the last decades conform to what we found for class differences.
\item In the long run, no clear connection between income inequality and social differences in mortality.
\end{itemize}

>>>>

To investigate the association between social class and mortality among adults and elderly from the 19th century to the present and how it is connected to socio-economic inequality in society.
This is analysed according to 
\begin{itemize}
\item gender
\item age group (working age vs retired)
\item *cause of death (cardiovascular, cancer)
\item *education 1990--2013)
\item *income 1986-2005)

@@

\begin{frame}{Economic inequality (figure from Roine and Waldenström 2008)}

@@

% \begin{frame}{Outline}
% 
% \begin{center}
% \begin{enumerate}
% 
% \item \emp{Data} management
% \item \emp{Modeling} considerations
% \item \emp{Analysis}
% \item \emp{Results}
% \item \emp{Implications} of the results
% \end{enumerate}
% 
% \end{center}
% 
% \end{frame}

@@

Remote area in northern Sweden, dominated by agriculture until well into the 20th century.

@@

\item Small town with rural hinterland, industrialization of the region during the 20th century.

@@

\begin{frame}{The sampling frame}
{Two separate datasets}

@@

\item[Sex]
\item[Marital status]
\item[Birth date] By quarter (15 Feb, May, Aug, Nov)
\item[Death date] By day (\emp{Socialstyrelsen}, \emp{Poplink})
\item[*Death cause] from \emp{Socialstyrelsen} and \emp{Poplink}. Adapted to ICD10
\item[Parish] Urban/Rural
\item[Social class] Elite, middle, worker. Household definition 1851--1950. Individual definition 1960--2013
\item[*Income] 1986--2005, grouped by quartiles.
\item[*Education] 1990--2013, three categories.
\item[Period]  1851--1875, \ldots,  2008--2013 (10 periods). No results presented for women 1961-70, 1971-80 due to missing information on class.
\item[Age group] Separate analyses for working active (40-65) vs retired (65-89) population
%\item[cohort] 1761--1771, \ldots, 1971--1974 (22 birth cohorts).

@@

% \begin{frame}[fragile]{Education 1990--2013}
% 
% <<educ>>=
% 
% educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
% educ$period <- factor(educ$period)
% educ <- educ[!is.na(educ$educ), ]
% levels(educ$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
% educ$educ <- relevel(educ$educ, ref = "middle")
% educ$civst <- relevel(educ$civst, ref = "married")
% fitm <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
%              data = educ[educ$sex == "male", ])
% fitf <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
%              data = educ[educ$sex == "female", ])
% cols <- c("darkgreen", "red", "black")
% @
% 
% <<ploteduc>>=
% oldpar <- par(mfrow = c(1, 2))
% 
% plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
% axis(1)
% axis(2, at = c(0, 0.5, 1.0), las = 1)
% abline(h = 0.5, lty = 3)
% box()
% legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
% 
% plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
% axis(1)
% axis(2, at = c(0, 0.5, 1.0), las = 1)
% abline(h = 0.5, lty = 3)
% box()
% legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
% 
% par(oldpar)
% @
% 
% \end{frame}

% \begin{frame}{Income 1986--2005}
% 
% <<femeduc>>=
% income <- lisaold[!is.na(lisaold$income), ]
% income$period <- factor(income$period)
% ##income <- income[!is.na(income$income), ]
% ##levels(income$income) <- c("low", "low", "middle", "middle", "middle", "high", "high")
% income$income <- relevel(income$income, ref = "low")
% income$civst <- relevel(income$civst, ref = "married")
% fitm <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
%              data = income[income$sex == "male", ])
% fitf <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
%              data = income[income$sex == "female", ])
% cols <- c("darkgreen", "red", "blue", "black")
% 
% @
% 
% <<plotincome>>=
% oldpar <- par(mfrow = c(1, 2))
% 
% plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
% axis(1)
% axis(2, at = c(0, 0.5, 1.0), las = 1)
% abline(h = 0.5, lty = 3)
% box()
% legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
% 
% plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
% axis(1)
% axis(2, at = c(0, 0.5, 1.0), las = 1)
% abline(h = 0.5, lty = 3)
% box()
% legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
% 
% par(oldpar)
% @
% 
% \end{frame}

@@

\begin{itemize}
\item \emp{Improved} survival characterised all social classes.
\item \emp{Late} appearance of the “modern” pattern of social differences in mortality among men
\item Distinct \emp{gender} differences in the development.
\item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
% \item No \emp{qualitative} but partly \emp{quantitative} differences between those in working age compared to the retired population.
% \item \emp{Educational} and \emp{income} differences during the last decades conform to what we found for class differences.
\item In the long run, no clear connection in the development of income inequality and social differences in mortality.
\item This suggests an important role of gender and class related expectations on health behaviour
\end{itemize}
